Global patterns of forest autotrophic carbon fluxes

### 2020-02-15 manuscript/manuscript draft.Rmd
rewrote first paragraph of intro

###Abstract
[**very rough start:**]
..... Here, we drew upon # records from the Global Forest Carbon Database (ForC), representing all major forest types and the nine most significant forest autotrophic carbon flux (FACF) variables, to comprehensively explore how C cycling in mature, undisturbed forests varies with latitude and climate on a global scale. We show that, across all FACF variables analyzed, C cycling decreases linearly with latitude --a finding that confirms multiple previous studies and contradicts the idea that productivity of temperate forests rivals that of tropical forests. The FACF variables increase in proportion to one another, with no differences in allocation detectable at this global scale, but differed in that latitude explained a lower proportion of variation among subsidiary fluxes (in particular, woody aboveground productivity and belowground net primary productivity, $BNPP$). Climate explained a significant proportion (#-#%) of variation in all C fluxes analyzed (less for subsidiary fluxes), with temperature variables in general and mean annual temperature ($MAT$) in particular being the best predictors of FACF on this global scale. While other climate variables (*e.g.*, **XX**) displayed significant correlation with FACF, none of them had significantly better explanatory power than $MAT$. The effects of temperature were modified by moisture availability, with reduced FACF under hot and dry conditions and sometimes under very high precipitaiton (especially for $BNPP$). FACF declined with temperature seasonality, but growing season length doesn't improve upon MAT as a predictor. Within the growing season, the influence of climate on C cycling is smaller but still significant for a number of carbon fluxes. These findings clarify the big picture of how FACF varies with latitude and climate on a global scale. As we enter a period of accelerating climatic change, understanding of the fundamental climatic controls on FACF sets a foundation for understanding patterns of change.

@@

(**revisit these first two paragraphs.**)

**Forests play a critical role in regulating atmospheric CO~2~ and climate** (**Bonan**). Approximately **##** Gt of CO~2~ cycle through Earth's forests each year **REF-KAT**), and their net C sequestration (~2.4 GtCyr^-1^) offsets roughly 30% of anthropogenic fossil fuel emissions [@pan_large_2011]. As atmospheric carbon dioxide levels continue to rise, driving climate change, there is increasing recognition that  protection and management of forest resources will play an important role in mitigating climate change (** REFS- e.g., IPCC 1.5C and land reports**). Carbon fixation, allocation, and metabolism by trees set the basis for energy and material flows in forest ecosystems and define their interactions with Earth's changing climate.

**Forests' autotrophic carbon flux (FACF, bbl:define) influences all organic matter stocks in ecosystems and is linked to cycling of energy, water, and nutrients** (**REFS**) [@piao_forest_2010]. Forest productivity sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance (**REFS**) and possibly diversity [@waide_relationship_1999] (**REFS**). However, our ability to draw general macroscopic conclusions regarding global variation in multiple FACF variables with respect to climate has been limited in that these analyses often mix forests that vary in stand age, disturbance history, and/or management status; do not always sufficiently parse related variables (*e.g.*, combining net primary productivity records with and without belowground components); and typically consider only one or a few variables at a time.

>>>>

###Abstract
[**very rough start:**]
Carbon fixation, allocation, and metabolism by trees set the basis for energy and material flows in forest ecosystems and define their interactions with Earth's changing climate.
Here, we drew upon # records from the Global Forest Carbon Database (ForC), representing all major forest types and the nine most significant forest autotrophic carbon flux (FACF) variables, to comprehensively explore how C cycling in mature, undisturbed forests varies with latitude and climate on a global scale. We show that, across all FACF variables analyzed, C cycling decreases linearly with latitude --a finding that confirms multiple previous studies and contradicts the idea that productivity of temperate forests rivals that of tropical forests. The FACF variables increase in proportion to one another, with no differences in allocation detectable at this global scale, but differed in that latitude explained a lower proportion of variation among subsidiary fluxes (in particular, woody aboveground productivity and belowground net primary productivity, $BNPP$). Climate explained a significant proportion (#-#%) of variation in all C fluxes analyzed (less for subsidiary fluxes), with temperature variables in general and mean annual temperature ($MAT$) in particular being the best predictors of FACF on this global scale. While other climate variables (*e.g.*, **XX**) displayed significant correlation with FACF, none of them had significantly better explanatory power than $MAT$. The effects of temperature were modified by moisture availability, with reduced FACF under hot and dry conditions and sometimes under very high precipitaiton (especially for $BNPP$). FACF declined with temperature seasonality, but growing season length doesn't improve upon MAT as a predictor. Within the growing season, the influence of climate on C cycling is smaller but still significant for a number of carbon fluxes. These findings clarify the big picture of how FACF varies with latitude and climate on a global scale. As we enter a period of accelerating climatic change, understanding of the fundamental climatic controls on FACF sets a foundation for understanding patterns of change.

@@

**Carbon cycling in forests worldwide is central to sustaining an estimated #% of Earth's biodiversity (*REF*), while strongly influencing atmospheric CO~2~ and climate** ([Bonan 2008](https://science.sciencemag.org/content/320/5882/1444)). Forests' autotrophic carbon flux (FACF)--that is, carbon fixation, allocation, and metabolism by trees and other primary producers--sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance (**REFS**) and possibly diversity [@waide_relationship_1999] (**REFS**). FACF influences all organic matter stocks in forest ecosystems and is linked to cycling of energy, water, and nutrients (**REFS**) [@piao_forest_2010]. Critically, FACF also defines forest interactions with Earth's changing climate. Approximately **##** Gt of CO~2~ cycle through Earth's forests each year **REF-KAT**), and in recent decades their net C sequestration (~2.4 GtCyr^-1^) offset roughly 30% of anthropogenic fossil fuel emissions [@pan_large_2011]. As atmospheric carbon dioxide levels continue to rise, driving climate change, forests will play a critical role in shaping the future of Earth's climate (** REFS**). However, our ability to draw general macroscopic conclusions regarding global variation in multiple FACF variables with respect to climate has been limited in that these analyses often mix forests that vary in stand age, disturbance history, and/or management status; do not always sufficiently parse related variables; and typically consider only one or a few variables at a time.
